---
title: "R: quick start"
---

This tutorial helps you check your R setup and learn how to run code and explore data. You’ll use these skills in later sections.

## Before you begin

1. **Open RStudio.** Double-click the RStudio icon (or launch it from Posit Cloud) and wait for the Console pane to show the R version. If the console displays an error, reinstall R before continuing.
2. **Create a script.** Select <kbd>File</kbd> → <kbd>New File</kbd> → <kbd>R Script</kbd>. This opens the Source pane where you can save your code as a reusable script. Save the file immediately using <kbd>Ctrl/Cmd</kbd> + <kbd>S</kbd> so you can keep track of your work.
3. **Run code.** Type `1 + 1` on a line in the Source pane and press <kbd>Ctrl/Cmd</kbd> + <kbd>Enter</kbd> (or use the **Run** button). The code should appear in the Console followed by the result. You can also type directly in the Console for quick checks.
4. **Check your working directory.** Run `getwd()` to confirm where files will be saved. If needed, use the Files pane to navigate to your project folder and click **Set As Working Directory**.

Once you are comfortable moving between the Source and Console panes, continue with installing and loading the packages you will use throughout the tutorials.


:::{.callout-tip}
There is a clipboard icon in the top right of each code block, you can use this to copy the code to the clipboard, which may be useful in some of the exercises. Copying does not execute the code — you’ll still need to paste and run it in your R console.
:::

## Installing and loading packages

For most of the work you do using R, you will need to use functions from packages. You install packages once, but you must load them each time you start R. For the tutorials on this site, you will need the following packages: `dplyr` for data manipulation, `ggplot2` for creating plots, and `broom` for summarising statistical models. These are installed using the `install.packages()` function.^[You can install multiple packages using one function call by using the `c()` operator to *chain* the packages together. Remember to use quote marks '"' around each package name.] If you are using a managed machine, you may need to request administrator rights or ask IT to complete this step for you.

```{r}
#| eval: false
install.packages(c("readr", "dplyr", "ggplot2", "broom"))
```


:::{.callout-note}
Running the `install.packages()` function will also install dependencies for the packages you have requested
:::

::: {.callout-tip}
You only need to install each package once per version of R, but you must use library() every time you start a new R session.
:::


You will then need to load each of these before using them, this must be done separately for each package and is best done at the top of each script. If `library()` reports an error, restart R (Session → Restart R) and try again to ensure your packages are available:

```{r}
#| message: false
library(readr) # For reading data
library(dplyr) # For data manipulation
library(ggplot2) # For creating plots
```

:::{.callout-note}
Anything after `#` is ignored by R. Use comments to explain what your code does or why you made certain choices.
:::


## Data

For more detail on importing data, see the R4DS chapter on [data import](https://r4ds.hadley.nz/data-import.html). The [RStudio data import cheatsheet](https://posit.co/resources/cheatsheets/read-data/) is also a handy reference when you are working with CSV, Excel, or database connections.

### Preloaded data

R comes with several datasets preloaded in the `datasets` package. These can be useful when you are learning new methods or to demonstrate things to others. For the tutorials on this site we will use the [`penguins` dataset](https://allisonhorst.github.io/palmerpenguins/reference/penguins.html), R (4.5.0+) includes the penguins dataset, originally from the palmerpenguins package. The first five rows are below:

```{r}
penguins |> 
  head(n = 5) # The head(n = 5) function shows the top five (n) rows of data
```


and the last four rows are below:

```{r}
penguins |> 
  tail(n = 4) # The tail(n = 4) function shows the bottom four (n) rows of data
```

::: {.callout-tip}
Use `glimpse(penguins)` to quickly see variable names, types, and example values.
:::

We can see that we have different types of data in each column. 

* `species`, `island`, and `sex` are factors
* `bill_len` and `bill_dep` are doubles (numeric variables with decimals)
* `flipper_len`, `body_mass`, and `year` are integers


:::{.callout-note}

The pipe operator `|>` essentially *passes* the object on the left to the right, this could be one one line

```
penguins |> head()
```

or it might be on multiple lines

```
penguins |> 
  head()
```

:::

### Load your own

Usually you will need to load your own data into R, this will need to be done each session. There are several ways of doing this, one of the most common ways to do so is using `read_csv()`

```{r}
#| eval: false

my_data <- read_csv("data/my_file.csv")

head(my_data)
```

:::{.callout-tip}
Make sure the file path is correct and that you’ve set your working directory to where the file is saved.
:::

## Explore dataset dimensions

Get a quick sense of how much data you have before diving into analysis.

::: {.panel-tabset}

### Example

Count the number of rows and columns in the `penguins` data.

```{r}
nrow(penguins)
ncol(penguins)
```

### Exercise

Work out how many distinct islands and years are represented. Replace the blanks with the correct column names and compute the summary with tidyverse verbs.

```{r}
#| eval: false

penguins |> 
  summarise(
    num_islands = n_distinct(____),
    num_years = n_distinct(____)
  )
```

### Solution

```{r}
#| code-fold: true
#| code-summary: "Show solution"

penguins |> 
  summarise(
    num_islands = n_distinct(island),
    num_years = n_distinct(year)
  )
```

:::

## Summarise numeric columns

Use basic summary functions to understand ranges of your numeric variables.

::: {.panel-tabset}

### Example

Compute the mean bill length while removing missing values.

```{r}
penguins |> 
  summarise(mean_bill_len = mean(bill_len, na.rm = TRUE))
```

### Exercise

Calculate the median flipper length and standard deviation of body mass, excluding missing values. Fill in the blanks with the appropriate column names.

```{r}
#| eval: false

penguins |> 
  summarise(
    median_flipper_len = median(____, ____ = ____),
    sd_body_mass = sd(____, ____ = ____)
  )
```

### Solution

```{r}
#| code-fold: true
#| code-summary: "Show solution"

penguins |> 
  summarise(
    median_flipper_len = median(flipper_len, na.rm = TRUE),
    sd_body_mass = sd(body_mass, na.rm = TRUE)
  )
```

:::

## Next chapter

Now that you can install, load, and explore data, you’re ready to start manipulating it with dplyr.
